---
title: "*Phytophthora sojae* Pathotype Data Analysis"
author: "Austin McCoy, Zachary Noel"
date: "December 19, 2018"
output:
    html_document: default
---
##### This analysis will provide Distribution of Susceptibilities, Distribution of complexities with statistics, Pathotype frequency distribution, as well as diversity indices for pathotypes.These scripts are meant to be a substitute for the Hagis spreadsheet previously used for *Phytophthroa sojae* pathotype analysis and provide the same necessary data as the Hagis sheet.
##### To start, your data should be in a similar format to that of the sample data file provided. Most importantly, Having columns labelled "Isolate"", "Rps"", and "perc.susc"" are critical for the code to work with minimal to no edits on the users part.

## Installing and loading Packages needed for analysis
```{r, warning=FALSE, message=FALSE}
# ipak will install and load the multiple r packages that are needed for this analysis
# Source: https://gist.github.com/stevenworthington/3178163

ipak <- function(pkg){
new.pkg <- pkg[!(pkg %in% installed.packages()[,"Package"])]
if (length(new.pkg)) 
    install.packages(new.pkg, dependencies = TRUE)
sapply(pkg, require, character.only = TRUE)
}

packages <- c("plyr", "ggplot2", "tidyverse", "plotrix", "stringr", "pander", "vegan")
ipak(packages)
```

```{r, message=FALSE}
# Reading in your data. Do not change "Pathotype.Data", just the file path to your data file, preferably in .csv format
# If there are any missing data points or values add ", na.strings = "" " after the name of your data set, but still within parentheses.
Pathotype.Data <- read_csv(YOUR FILE PATHWAY HERE)
```


## Section 1: Distribution of Susceptibilities
```{r}
#This function with calculate the distribution of susceptibilities by Rps gene. You can change your susceptible cutoff value at the bottom of this chunk labelled "Distribution_of_Susceptibilities(60)" to whatevevr percentage you prefer for susceptible or resistant reactions
Distribution_of_Susceptibilities = function(susceptibility_cutoff) {
# if else for resistant or susceptible reaction. This will mark susceptible reactions with a "1" in a new column labelled "Susceptible.1" to then be used in later analysis.

Pathotype.Data$Susceptible.1 <- ifelse(Pathotype.Data$'perc.susc' >= susceptibility_cutoff, 1, 0)

## summary by rps gene to tally. This code takes the "Susceptible.1" column and summarises it by Rps gene for your total Isolates pathogenic on each gene. Likewise "Isolate_N" is calculated given the unique Isolate names to find the total number of isolates within your data set. "Percent_isolates_pathogenic" is then found for each gene, showing the percentage of isolates that are pathogenic on tested genes. "Rps.Gene.Summary" will return these values.

Rps.Gene.Summary <- ddply(Pathotype.Data, c("Rps"), summarise, N=sum(Susceptible.1))
Pathotype.Data$Isolate <- factor(Pathotype.Data$Isolate)
Isolate_n <- length(levels(Pathotype.Data$Isolate))
Rps.Gene.Summary$percent_isolates_pathogenic <- ((Rps.Gene.Summary$N)/Isolate_n)*100
Rps.Gene.Summary

# visualization of data, using "Rps.Gene.Summary" to showing the percentage of isolate pathogenic on each gene tested. Scripts below can be edited for visual appeal.

Visualization_of_Susceptibilities <- ggplot(data=Rps.Gene.Summary,aes(x = Rps , y = percent_isolates_pathogenic)) +
  stat_summary(fun.y=mean,position=position_dodge(width=0.95),geom="bar") +
    theme_classic() +
    ggtitle("Percentage of Isolates Pathogenic on Rps genes") + 
    theme(axis.text.x = element_text(size = 12, face = "bold", angle=45, hjust=1, family = "serif"),
        axis.text.y = element_text(size = 15, face = "bold", family = "serif"),
        axis.title.x = element_text(size = 25, face = "bold", family = "serif"),
        axis.title.y = element_text(size = 20, face = "bold", family = "serif"),
        axis.line.x = element_line(colour = 'black', size=0.5, linetype='solid'),
        axis.line.y = element_line(colour = 'black', size=0.5, linetype='solid'),
        legend.text = element_text(size = 10, face = "bold", family = "serif"),
        legend.key = element_blank(),
        legend.title = element_text(size = 10, face="bold", family = "serif"),
        legend.position = "right",
        strip.text.x = element_text(size = 25, face = "bold", family = "serif"),
        title = element_text(size = 10, family = "serif")) +
  labs(list(y = expression(bold("Percent of Isolates"))),
       x = expression(bold("Rps Gene"))) 
final <- list(Rps.Gene.Summary, Visualization_of_Susceptibilities)
return(final)
}
#The value in "Distribution_of_Susceptibilities(60)" (in this case, 60), sets the cutoff for susceptible reactions. For example, currently all genes with 60% or more of the plants rated susceptible will return a "1" in previous scripts (see line 30).
Distribution_of_Susceptibilities(60)
```

## Section 2: Distribution of Complexities
```{r}
#Similar to the previous chunk, the "Susceptibility_cutoff" can be changed at the bottom of this chunk
Distribution_of_Complexities = function(susceptibility_cutoff) {

# The susceptible control is removed from all isolates in the data set so that it will not impact complexity calculations and a new data set is made that does not contain susceptible controls. Thus, complexities can be from 0 to 13 (13 genes tested) for this data set. This can be changed later on if more, or less, genes are being tested.
# for this to work your susceptible control must be labelled "susceptible" under the Rps column of your data set. you can change "susceptible" to "rps/rps" or whatever you have it labelled as and it should work.
  remove_controls <- subset(Pathotype.Data, Rps != "susceptible")

  Pathotype.Data$Susceptible.1 <- ifelse(Pathotype.Data$'perc.susc' >= susceptibility_cutoff, 1, 0)

# The new data set "remove_controls" has the susceptible calculations performed again for, just like in the previous chunk, only without the susceptible control.
remove_controls$Susceptible.1 <- ifelse(remove_controls$'perc.susc' >= susceptibility_cutoff, 1, 0)


Rps.Gene.Summary <- ddply(Pathotype.Data, c("Rps"), summarise, N=sum(Susceptible.1))
Pathotype.Data$Isolate <- factor(Pathotype.Data$Isolate)
Isolate_n <- length(levels(Pathotype.Data$Isolate))
#Rps.Gene.Summary$percent_isolates_pathogenic <- ((Rps.Gene.Summary$N)/Isolate_n)*100
#Rps.Gene.Summary

#Individual Isolate Complexities as calculated by grouping by Isolate and then summarising the number of"1"'s for each Isolate in the "Susceptible.1" Column.
Ind_complexities <- remove_controls %>%
  group_by(Isolate) %>%
  summarise(N=sum(Susceptible.1))


# Frequency for each complexity (%). 
#Percent frequency is calculated by taking the individual complexity of each Isolate and grouping all Isolates by their complexity (Ind_complexities$Complexity[Ind_complexities$N==X]). The function legnth() is then used to identify how many of each complexity there are in the data set (from 0 to 13). This number is then taken and divided by the total number of isolates in the dataset (which was calculated in previous scripts and is identified as "Isolate_n" here) then multiplied by 100 for final percentages.
# If you are using less than 13 genes, you can edit the script by placing "#" before all "Freq_comp_X" that are above your number of genes tested. If testing more than 13 genes, you will need to simply copy a line of this code and add as many more linesas necessary, making sure to change the "N==X" to the correct complexity (i.e. 14, 15, 16, etc)
# Changing the script here will require edits on scripts down the line, but I will try and annotate where those changes would need to be made, if necessary

Ind_complexities$Complexity <- as.character(Ind_complexities$N) #adds a column telling the complexity of each individual isolates in the dataset
# Determining the frequency of each complexity (0:13) within the data set
freq_comp_0 <- (length(Ind_complexities$Complexity[Ind_complexities$N==0])/Isolate_n)*100
freq_comp_1 <- (length(Ind_complexities$Complexity[Ind_complexities$N==1])/Isolate_n)*100
freq_comp_2 <- (length(Ind_complexities$Complexity[Ind_complexities$N==2])/Isolate_n)*100
freq_comp_3 <- (length(Ind_complexities$Complexity[Ind_complexities$N==3])/Isolate_n)*100
freq_comp_4 <- (length(Ind_complexities$Complexity[Ind_complexities$N==4])/Isolate_n)*100
freq_comp_5 <- (length(Ind_complexities$Complexity[Ind_complexities$N==5])/Isolate_n)*100
freq_comp_6 <- (length(Ind_complexities$Complexity[Ind_complexities$N==6])/Isolate_n)*100
freq_comp_7 <- (length(Ind_complexities$Complexity[Ind_complexities$N==7])/Isolate_n)*100
freq_comp_8 <- (length(Ind_complexities$Complexity[Ind_complexities$N==8])/Isolate_n)*100
freq_comp_9 <- (length(Ind_complexities$Complexity[Ind_complexities$N==9])/Isolate_n)*100
freq_comp_10 <- (length(Ind_complexities$Complexity[Ind_complexities$N==10])/Isolate_n)*100
freq_comp_11 <- (length(Ind_complexities$Complexity[Ind_complexities$N==11])/Isolate_n)*100
freq_comp_12 <- (length(Ind_complexities$Complexity[Ind_complexities$N==12])/Isolate_n)*100
freq_comp_13 <- (length(Ind_complexities$Complexity[Ind_complexities$N==13])/Isolate_n)*100

# Making another dataset which only contains the frequncey of each complexity, for use in data visualization
Freq_of_complex <- rbind.data.frame(freq_comp_0, freq_comp_1, freq_comp_2, freq_comp_3, freq_comp_4, freq_comp_5, freq_comp_6, freq_comp_7, freq_comp_8, freq_comp_9, freq_comp_10, freq_comp_11, freq_comp_12, freq_comp_13)
colnames(Freq_of_complex) = "Frequency_of_Complexities"
# adds another column to the data set so you can easily tell the frequency of each complexity in your data.
# This will need to be altered should you be testing a different number of genes. Change the "13" to you maximum complexity possible, without controls
Freq_of_complex$complexities <- (0:13)

## Distribution of complexity (counts)
# This operates in the same manner as the frequency data, but will only show how many (not percentage) isolates are included in each complexity
# Changes made to frequency data for more, or less, genes will also need to be made here.
Distr_comp_0 <- length(Ind_complexities$Complexity[Ind_complexities$N==0])
Distr_comp_1 <- length(Ind_complexities$Complexity[Ind_complexities$N==1])
Distr_comp_2 <- length(Ind_complexities$Complexity[Ind_complexities$N==2])
Distr_comp_3 <- length(Ind_complexities$Complexity[Ind_complexities$N==3])
Distr_comp_4 <- length(Ind_complexities$Complexity[Ind_complexities$N==4])
Distr_comp_5 <- length(Ind_complexities$Complexity[Ind_complexities$N==5])
Distr_comp_6 <- length(Ind_complexities$Complexity[Ind_complexities$N==6])
Distr_comp_7 <- length(Ind_complexities$Complexity[Ind_complexities$N==7])
Distr_comp_8 <- length(Ind_complexities$Complexity[Ind_complexities$N==8])
Distr_comp_9 <- length(Ind_complexities$Complexity[Ind_complexities$N==9])
Distr_comp_10 <- length(Ind_complexities$Complexity[Ind_complexities$N==10])
Distr_comp_11 <- length(Ind_complexities$Complexity[Ind_complexities$N==11])
Distr_comp_12 <- length(Ind_complexities$Complexity[Ind_complexities$N==12])
Distr_comp_13 <- length(Ind_complexities$Complexity[Ind_complexities$N==13])

# Producing another dataset for the count/distribution of complexities within the data, same as frequency
Distr_of_complex <- rbind.data.frame(Distr_comp_0, Distr_comp_1, Distr_comp_2, Distr_comp_3, Distr_comp_4, Distr_comp_5, Distr_comp_6, Distr_comp_7, Distr_comp_8, Distr_comp_9, Distr_comp_10, Distr_comp_11, Distr_comp_12, Distr_comp_13)
colnames(Distr_of_complex) = "Distribution_of_Complexities"
# This will need to be altered should you be testing a different number of genes. Change the "13" to you maximum complexity possible, without controls
Distr_of_complex$complexities <- (0:13)

## Visualization of Frequency (%)
Visualization_of_complexities <- ggplot(data=Freq_of_complex,aes(x = complexities , y = Frequency_of_Complexities)) +
    geom_bar(stat="identity") +
  scale_x_continuous(name = "Complexity of Pathotype", breaks = seq(0,13,1)) +
  scale_y_continuous(breaks = seq(0,20,2)) +
   theme_classic() +
    ggtitle("Percentage of Pathotype Complexities") + 
    theme(axis.text.x = element_text(size = 12, face = "bold", angle=0, family = "serif"),
        axis.text.y = element_text(size = 15, face = "bold", family = "serif"),
        axis.title.x = element_text(size = 25, face = "bold", family = "serif"),
        axis.title.y = element_text(size = 20, face = "bold", family = "serif"),
        axis.line.x = element_line(colour = 'black', size=0.5, linetype='solid'),
        axis.line.y = element_line(colour = 'black', size=0.5, linetype='solid'),
        legend.text = element_text(size = 10, face = "bold", family = "serif"),
        legend.key = element_blank(),
        legend.title = element_text(size = 10, face="bold", family = "serif"),
        legend.position = "right",
        strip.text.x = element_text(size = 25, face = "bold", family = "serif"),
        title = element_text(size = 10, family = "serif")) +
  labs(list(y = expression(bold("Frequency Isolates per complexity"))),
       x = expression(bold("Complexity of Pathotype")))

## Visualization of Distribution (N per complexity)
Visualization_of_Distribution <- ggplot(data=Distr_of_complex,aes(x = complexities , y = Distribution_of_Complexities)) +
    geom_bar(stat="identity") +
  scale_x_continuous(name = "Complexity of Pathotype", breaks = seq(0,13,1)) +
  scale_y_continuous(breaks = seq(0,20,2)) +
   theme_classic() +
    ggtitle("Number of Isolates in each Pathotype Complexity") + 
    theme(axis.text.x = element_text(size = 12, face = "bold", angle=0, family = "serif"),
        axis.text.y = element_text(size = 15, face = "bold", family = "serif"),
        axis.title.x = element_text(size = 25, face = "bold", family = "serif"),
        axis.title.y = element_text(size = 20, face = "bold", family = "serif"),
        axis.line.x = element_line(colour = 'black', size=0.5, linetype='solid'),
        axis.line.y = element_line(colour = 'black', size=0.5, linetype='solid'),
        legend.text = element_text(size = 10, face = "bold", family = "serif"),
        legend.key = element_blank(),
        legend.title = element_text(size = 10, face="bold", family = "serif"),
        legend.position = "right",
        strip.text.x = element_text(size = 25, face = "bold", family = "serif"),
        title = element_text(size = 10, family = "serif")) +
  labs(list(y = expression(bold("Distribution of Complexities"))),
       x = expression(bold("Complexity of Pathotype")))

## summary statistic (Distribution). Mean, Standard deviation and Standard error of complexities for your data
# this will print as 3 = mean, 4 = standard deviation and 5 = standard error
comp_mean <- mean(Ind_complexities$N)
comp_sd <- sd(Ind_complexities$N)
comp_stderr <- std.error(Ind_complexities$N)


# This code will return all of the data which we just made once this chunk has been run, no need to go in and run everything one at a time. Simply run the chunk in an Rmd file and the output will appear either in the console (statistics) or in the plot visualization (frequency and distribution of complexities)
final_complexities <- list(Freq_of_complex, Distr_of_complex, comp_mean, comp_sd, comp_stderr, Visualization_of_complexities, Visualization_of_Distribution)

return(final_complexities)
}
#Again, you can change your susceptible cutoff value here for your dataset
Distribution_of_Complexities(60)

```

## Section 3: Pathotype Frequency Distribution
```{r, message = FALSE}
# unlike previous scripts, we need to define the cutoff value here, instead of at the bottom of the chunk. (I will look into this and figure out why I get a weird output when using a function...)
cutoff_value <- 60

# same as previous scripts
remove_controls <- subset(Pathotype.Data, Rps != "susceptible")
remove_controls$Susceptible.1 <- ifelse(remove_controls$'perc.susc' >= cutoff_value, 1, 0)

# Removal of resistant reactions from the data set, leaving only susceptible reactions (pathotype)
Remove_resistance <- subset(remove_controls, Susceptible.1 != 0) %>%
  transform(Rps = str_replace(Rps, "Rps ", "")) # this line takes the "Rps" out of my data set leaving only the gene number, as you would see in a publication. You may not need this line for yours...

#Individual Isolate Complexities
# using our data set that now only has susceptible reactions, the actual pathotype for each individual Isolate is now displayed. Print "Ind_pathotypes" to take a look!
Ind_pathotypes <- Remove_resistance %>%
  group_by(Isolate) %>%
  nest() %>%
  mutate(Pathotype = map(data, ~ toString(.$Rps)))  %>%
  unnest(Pathotype) %>%
  select(Isolate, Pathotype)

# Identifying the frequency at which each Pathotype is found in the data set
#  Isolate needs to be a character vector for this to work, this line of code takes care of that
Ind_pathotypes$Isolate <- as.character(Ind_pathotypes$Isolate)
# The frequency at which each pathotype is found within the dataset is performed here. It can be confusing to look at, but we will clean it up in the next step. For now, each isolates pathotype will have a column next to it, showing how often that pathotype is in the dataset.
Pathotype_Freq <- within(Ind_pathotypes, { count <- ave(Isolate, Pathotype, FUN=function(Pathotype) length(unique(Pathotype)))})

#Final Chart for visualizing unique pathotype distributions
# this script takes out only the unique pathotypes and the count at which they are found in the data set to be used in the next graphic
Pathotype_Freq_Distribution <- Pathotype_Freq %>%
    select(count, Pathotype) %>%
 distinct(Pathotype, .keep_all = TRUE)
# table showing only unique pathotypes and their frequency within the dataset
pander::pander(Pathotype_Freq_Distribution)
```

## Section 4. Diversity index for Pathotypes
```{r}
# Diversity indices used to investigate pathotype divversity within and between states are shown below. In Version 1 of this document, only code for analyzing a single state at a time is shown. In the future, scripts could be produced so that multiple states could be analyzed at once, independently of each other. Therefore, if analyzing multiple states pathotype data, each state must be analyzed from its own .csv document.

# changes "$count" to a numeric value
Pathotype_Freq_Distribution$count <- as.numeric(Pathotype_Freq_Distribution$count)
# switches the column order of "Pathotype" and "count" for the analysis
Pathotype_Freq_Distribution <- Pathotype_Freq_Distribution[,c("Pathotype","count")]
#Changes Isolate to a factor
Pathotype.Data$Isolate <- factor(Pathotype.Data$Isolate)
# determines the number of isolates within the data
Isolate_n <- length(levels(Pathotype.Data$Isolate))

# renaming "Isolate_n" to "Number_of_isolates"
Number_of_isolates <- Isolate_n
# Determining the number of unique pathotypes for this analysis
Number_of_pathotypes <- specnumber(Pathotype_Freq_Distribution$count)


Simple <- Number_of_pathotypes/ Number_of_isolates
Gleason <- (Number_of_pathotypes - 1)/log(Number_of_isolates)
Shannon <- diversity(Pathotype_Freq_Distribution[-1], index="shannon")
Simpson <- diversity(Pathotype_Freq_Distribution[-1], index="simpson")
Evenness <- Shannon/ log(Number_of_pathotypes)

Number_of_isolates
Number_of_pathotypes

# Simple diversity will show the proportion of unique pathotypes to total isolates. As the values gets closer to 1, there is greater diversity in pathoypes within the population.
Simple

# an alternate version of Simple diversity index. This index is less sensitive to sample size than the simple index.
Gleason

# Shannon diversity index is typically between 1.5 and 3.5. As richness and evenness of the population increase, so does the Shannon index value
Shannon

# Simpsom diversity index values range from 0 to 1. 1 represents high diversity and 0 represents no diversity.
Simpson

# evenness ranges from 0 to 1. As the Eveness value approaches 1, there is a more evene distribution of each pathoypes frequency within the population.
Evenness
```


### Recommendations are always appreciated!!
#### Chilvers Lab
#### Michigan State University
#### East Lansing, MI
