---
title: "hagis: Tools for Analysis of Phytophthora Pathotype Complexities, Distributions and Diversity"
author: "Austin McCoy and Zachary Noel"
date: "`r Sys.Date()`"
output: 
    rmarkdown::html_vignette:
        toc: true
vignette: >
  %\VignetteIndexEntry{hagis: Tools for Analysis of Phytophthora Pathotype Complexities, Distributions and Diversity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
library(hagis)
```

## Getting Started With *hagis*

First you'll want to load in your data set, for right now let's use a practice data set made for the *hagis* package and name it `Pathotype.Data`:

```{r load_data}
Pathotype.Data <- system.file("extdata", "practice_data_set.csv", package = "hagis")
Pathotype.Data <- read.csv(Pathotype.Data, stringsAsFactors = FALSE)
head(Pathotype.Data)
```

We see in the `Rps` column that each gene is prepended with "Rps". We can remove this to make the graphs cleaner and report the genes in tables as we would in a manuscript.

```{r remove-Rps}
Pathotype.Data$Rps <- gsub("Rps ", "", Pathotype.Data$Rps)
head(Pathotype.Data)

```

This practice data set contains 21 isolates' virulence data on a set of 14 differential soybean cultivars. This package uses the percentage of susceptible, inoculated, plants to determine effective resistance genes, pathotype diversity and frequency, as well as individual isolates pathotypes.

***

## Function Arguments Used in *hagis*

We have strived to make *hagis* as intuitive to use as possible. Part of that means that we have used the same arguments for the three main functions, `summarize_rps()`, `calculate_complexities()` and `calculate_diversity()`. Each of these functions take the same arguments:

* `x` followed by your data set name allows for the function to identify where it will be pulling these columns (and their associated row values) from to use (_i.e._ your data collection Excel sheet)

* `sample` specifies the column header for the column which identifies the isolates tested

* `perc_susc` specifies the column header for the column which identifies the percent susceptible plants for each gene

* `Rps` specifies the column header for the column which identifies the genes tested

* `control` specifies the value used in the `Rps` column to denote a susceptible control

* `cutoff` this value sets the cutoff for susceptible reactions. For example, `cutoff = 60` means that all genes with 60% or more of the plants rated susceptible will be treated as susceptible. You can change this to whatever percentage you require for your study.

***

## Determination of Effective Resistance Genes

Below is an example of tables and graphics that can be produced using the `summarize_rps()` function to identify effective resistance genes tested against the sampled *Phytophthora sojae* population. 

The `summarize_rps()` function allows you to produce a detailed table showing the number of isolates each gene is susceptible to (`N_susc`), as well as offering a percentage of the isolates tested which are pathogenic on each gene (`percent_pathogenic`).

```{r, echo=TRUE}
Rps.summary <- summarize_rps(x = Pathotype.Data,
              cutoff = 60,
              control = "susceptible",
              sample = "Isolate",
              Rps = "Rps",
              perc_susc = "perc.susc")

Rps.summary
```

Using the _pander_ library we can make the table much more attractive in Markdown.

```{r pander-print-Rps, echo=TRUE}
library(pander)

pander(Rps.summary)
```

### Plotting Rps Summary Data

*hagis* also provides functions to quickly graph your data using `ggplot2`.

```{r autoplot-summary, echo=TRUE}
autoplot(Rps.summary, type = "percentage")
```

You might note that the x-axis are on the left, not the bottom. This is because the Rps gene names are long enough that they need to be rotated if on the bottom. This way it is much easier to read the plot axis and interpret the plots. For more on plotting data from *hagis* see the "[Advanced Plotting](#AdvancedPlotting)" section.

***

## Pathotype Complexities

Pathotype frequency, distribution as well as statistics such as mean pathotype complexity can be calculated using the `calculate_complexities()` function. This function will return
a `list()` of two `data.table()` objects, `grouped_complexities` and `individual_complexities`.

```{r complexities, echo=TRUE, message=FALSE, warning=FALSE}
complexities <- calculate_complexities(x = Pathotype.Data, 
                                       cutoff = 60, 
                                       control = "susceptible", 
                                       sample = "Isolate", 
                                       Rps = "Rps", 
                                       perc_susc = "perc.susc")

complexities
```

Once again, using _pander_ we can make these tables much more attractive in Markdown. Since `complexities` is a `list()` object, we can refer to each object directly by name and print them as follows.

```{r pander-print-complexities}
pander(complexities$grouped_complexities)

pander(complexities$indvidual_complexities)
```

Using `summary()` will return the mean, standard error (se) and standard deviation (sd) for pathotypes of a complexities object.

```{r summary-complexities}
pander(summary(complexities))
```

### Plotting Complexities Data

Two functions are provided to plot the complexities depending on your needs. If you need the frequency, use `autoplot(complexties, type = "frequency")`, or if you desire the distribution `autoplot(complexities, type = "distribution")`. Both return the same graph, only the y-axis change; percent for frequency and n for distribution.

```{r complexities-plot}
autoplot(complexities, type = "percentage")

autoplot(complexities, type = "count")
```

***

## Diversity Indices, Frequency of Unique Pathotypes and Individual Isolate Pathotypes

Diversity indices are extremely useful when trying to identify differences between two populations. Here, pathotype diversities are calculated for the sample population using the `calculate_diversities()` function. Likewise, individual isolates' pathotypes, number of isolates used in the study, number of pathotypes within the study are calculated. 

```{r calculate-diversities, echo=TRUE}
diversity <- calculate_diversities(x = Pathotype.Data,
                                   cutoff = 60,
                                   control = "susceptible",
                                   sample = "Isolate",
                                   Rps = "Rps",
                                   perc_susc = "perc.susc")
diversity
```

Or using `pander` for reporting, a nice table is generated.

```{r diversity-pander}
pander(diversity)
```

### Table of Diversities

To generate a table of diversities, use `diversities_table()`. *hagis* will automatically create a `pander` object for you. This is because it is much easier to read the resulting table in the console than the raw `data.frame` and insert into reports.

```{r diversities-table}
diversities_table(diversity)
```

To generate a table of individual pathotypes, use `individual_pathotypes()`. Here again,
*hagis* provides a `pander` object for ease of use.

### Table of Individual Pathotypes

```{r individual-pathotypes}
individual_pathotypes(diversity)
```

# Advanced Plotting {#AdvancedPlotting}

## Using *hagis* Autoplot Objects

Since *hagis* uses `ggplot2` to generate its plots, you can easily theme the outputs using common `ggplot2` themes and other options.

```{r set-up-adv.plot}
library(ggplot2)

Rps.plot <- autoplot(Rps.summary, type = "percentage")

Rps.plot
```

### Changing the ggplot2 Theme

Use *ggplot2*'s `theme_minimal()` theme.

```{r change-autoplot-theme}
Rps.plot <- Rps.plot +
  theme_minimal()

Rps.plot
```

### Changing the Font

Set the font to be a bold-face serif family font.

```{r change-autoplot-font}
Rps.plot <- Rps.plot +
  theme(
    text = element_text(
      face = "bold",
      family = "serif"
    ))

Rps.plot
```

### Make a Horizontal Plot

Set the font to be a bold-face serif family font a

```{r horizontal-plot}
Rps.plot <- Rps.plot +
  coord_flip()

Rps.plot
```

### Use Colours in Autoplot Objects

You can use named, *e.g.* "red", "yellow", "blue" colors in R or you can use custom hexadecimal color codes. Illustrated below is using MSU Green, #18453b.

```{r use-colours}
autoplot(Rps.summary, type = "percentage", color = "#18453b")
```
